%*******************************************************
% Acknowledgments
%*******************************************************
\pdfbookmark[1]{Acknowledgements}{acknowledgements}
\manualmark
\markboth{\spacedlowsmallcaps{Acknowledgements}}{\spacedlowsmallcaps{Acknowledgements}}

\bigskip

\begingroup
\let\clearpage\relax
\let\cleardoublepage\relax
\let\cleardoublepage\relax
\chapter*{Acknowledgements}

\def\thanks#1{%
\begingroup
\leftskip1em
\noindent #1
\par
\endgroup
}

My parents---for raising my curious mind.

My brothers---I learned more from them than they realize.

\thanks{All my teachers---whom I spent many years annoying but remember every last one of. I hope this thesis makes up for at least some of the trouble.}

\thanks{Vladimir---for giving me the chance at a PhD at ETH, my home university. He asked me the right questions many times.}

\thanks{Everyone in the Deupi and Schertler group(s) deserves thanks for welcoming me into the lab, teaching me to use a pipette, sharing coffee, and patiently listening to my ramblings about AI, language models, and whatever else I was excited about that week.}

\thanks{I thank the members of the High-Performance Computing and Emerging Technologies(HPCE) group at the Paul Scherrer Institute for technical support.}

\thanks{And finally, Xavi---for his mentorship and for giving me remarkable freedom in my research from an early stage. He let me explore while always keeping me on track. He has the best explanations and a radar for whenever I was hand-waving. He would catch it immediately and help me get it right.}

\thanks{Eleven years at ETH brought me into contact with remarkable students, researchers and professors. The Paul Scherrer Institute and its staff provided an environment where computational and experimental work met daily.}

\thanks{This work was supported by the Swiss National Science Foundation (grant no.\ 192780) and by the European Union's Horizon 2020 research and innovation programme (grant agreement no. 951644)}

\thanks{I acknowledge Claude Opus-4.5 and Opus-4.6 (Anthropic) for editing assistance in preparing this thesis. Claude was also used as the brain (LLM) in the ProtOS-MCP example (\autoref{ch:protos-mcp}). The scientific work, analysis, and conclusions are entirely my own.}
\endgroup
